\section{Sequences}

\begin{defn}[Sequence, tuple, ordered pair]\label{defn:sequence}\label{defn:tuple}
A \defterm{sequence} is a list of objects in some order. We usually denote a sequence by writing the list of objects surrounded by parentheses. Order doesn't matter in a set, but in a sequence it does. Sequences may be finite or infinite. Finite sequences are often called \defterm{tuples}. A sequence with $k$ elements is a \defterm{$\bm k$-tuple}. Thus, $(7, 21, 42)$ is a 3-tuple. A 2-tuple is also called an \defterm{ordered pair}.
\end{defn}

\begin{discussion}
Sets and sequences often appear as elements of other sets and sequences. 
\end{discussion}

\begin{defn}[Power set]
The \defterm{power set} of a set $A$, written $\bm{\powerset{A}}$, is the set of all subsets of $A$.
\end{defn}

\begin{exer}
What are the power sets of each of the following?
\begin{itemize}
\item $\emptyset$
\item $\setdef{\emptyset}$
\item $\setdef{a, b, c}$
\end{itemize}
\end{exer}

\begin{defn}[Cartesian product]
The \defterm{Cartesian product} or \defterm{cross product} of two sets, $A$ and $B$, written $\bm{\setproduct{A}{B}}$, is the set of all ordered pairs where the first element is a member of $A$ and the second element is a member of $B$. We can take the Cartesian product of $k$ sets, $A_1, A_2, \ldots, A_k$, which is the set consisting of all $k$-tuples $(a_1, a_2, \ldots, a_k)$ where $a_i \in A_i$.

When we have the Cartesian product of a set $A$ some number of times with itself, we'll use the shorthand:
\[
A^k \ =\  \overbrace{A \times A \times \cdots \times A}^k
\]

\end{defn}

\begin{exer}
Fill in the following definitions of power set and Cartesian product using set-builder notation.
\begin{itemize}
\item $\powerset{A} = \setbuild{S}{\line(1,0){250}}$
\item $\setproduct{A}{B} = \setbuild{(a, b)}{\line(1,0){250}}$
\end{itemize}
\end{exer}

\begin{exer}
If $A = \setdef{1, 2}$ and $B = \setdef{x ,y, z}$, what is $\setproduct{A}{\setproduct{B}{A}}$?
\end{exer}

\begin{exer}
Describe $\naturals^2$. ($\naturals$ is the set of natural numbers.)
\end{exer}

\begin{exer}
If $|A| = m$ and $|B| = n$, then what is $|\setproduct{A}{B}|$?
\end{exer}
